Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-03-12 supplement.tex
a final read-through for small errors

DNA samples were collected from three Nama communities in the Richtersveld
region of South Africa, which borders southern Namibia, in 2012. A community
guide was present during each interview and facilitated consent in Afrikaans or
Nama. Written consent was recorded per our IRB protocol with human subjects
approval from Stanford University (Protocol \#13829), Stellenbosch University
(N11/07/210) and later maintained via SUNY Stony Brook (Protocol \#727494).
Saliva samples were extracted from Oragene [OGR-500] kits at Stellenbosch
University. Results stemming from genetic analyses have been communicated in
2015, 2019, 2021 via community presentations, a radio interview and to
representatives of the Richtersveld National Park.

@@

From the merged dataset of the African Diversity Reference Panel (ADRP,
citet{}) and the \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data,
we subsampled to the populations included in this study. From the 1000 Genomes
data, these included all populations in the African super-population (ESN, GWD,
LWK, MSL, and YRI) and four non-African populations (CEU, GBR, CHB, and PJL).
From the ADRP, these included the Amhara, Bakiga, Gumuz, Nama, Oromo, Somali,
Nama, and Zulu. This larger set of populations were used in the PCA and
\texttt{Admixture} analyses. The \texttt{Relate} analysis used a slightly
smaller set, discluding the Bakiga, Somali, and Zulu samples.

@@

All analyses presented in this work focus on biallelic single nucleotide
polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the
\texttt{moments-LD} analysis, we focused on intergenic regions because these
appear less affected by natural selection compared to both synonymous and
nonsynonymous variation \citep{Ragsdale2018-dd}. To enable comparison with
Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
had less than 100 contiguous base pairs.

@@

We used principal component analysis (PCA) from \texttt{scikit-learn}
\citep{Pedregosa2011-ke} and \texttt{ADMIXTURE} version 1.3.0
\citep{Alexander2011-aq} to summarize genetic diversity among a broader set of
1000 Genomes and ADRP populations (Figure~\ref{fig:diversity} in the main text). Data
filtering was identical for both analyses. We retained biallelic SNPs with no
missing data and minor allele frequency greater that 5\%. We then thinned the
remaining sites to remove SNPs in high linkage disequilibrium with one another.
We used the \texttt{locate\_unlinked} function from \texttt{scikit-allel}
version 1.2.1 \citep{Miles2021-yq} to identify SNPs in low linkage
disequilibrium, using a threshold of $r^2=0.1$, a window size of 100 SNPs, and
a step size of 20 SNPs, removing the remaining SNPs. This process was iterated
five times for each chromosome, resulting in 327,656 roughly unlinked SNPs used
in the PCA and \texttt{ADMIXTURE} analyses.

@@

Given two biallelic loci, with alleles $A$ and $a$ and the left locus and
alleles $B$ and $b$ at the right locus, the standard covariance measure of LD is
$D=p_{AB}p_{ab}-p_{Ab}p_{aB}$, where $p_{AB}$ is the frequency of $AB$
haplotypes in a population (and thus the probability of drawing an $AB$
haplotype in a random sample of that population). \citet{Hill1968-vu} solve for
the expectation of $D^2$ using a system of equations that includes $\E[Dz] =
\E[D(1-2p_A)(1-2p_B)]$ and $\E[\pi_2] = \E[p_A(1-p_A)p_B(1-p_B)]$, where $p_A$
and $p_B$ are the frequencies of $A$ and $B$ at the left and right loci,
respectively. This system also requires the expected pairwise diversity (or
expected heterozygosity, assuming random mating), denoted $\E[H] =
\E[2p_A(1-p_A)] = \E[2p_B(1-pB)]$, assuming equal mutation rates at the two
loci.

@@

$\sigma_{Dz}$ has been shown to be sensitive to deep population structure and
archaic hominin admixture \citep{Ragsdale2019-nt}, and this statistic is closely
related to $S^*$ statistics used to scan for introgressed haplotypes
\citep{Plagnol2006-lt}. Pairwise diversity statistics $H_{i,j}$ have also been
widely used in demographic inference involving ancient DNA and many samples, as
$f_2$, $f_3$ and $f_4$ statistics can be expressed as linear combinations of
$H_{i,j}$. $f$-statistics form the basis of admixture graph analysis
\citep{Lipson2020-cq}. Therefore, the set of statistics used here encompass
multiple features of genetic data that have been used to infer models of
archaic hominin admixture and population structure involving many populations.

@@

For numerical convenience, observed genetic variants were binned by
recombination distances. We assessed the robustness of the statistics to errors
in the recombination maps by considering two different recombination maps, the
OMNI YRI and HapMapII
\citep{1000_Genomes_Project_Consortium2015-zq,International_HapMap_Consortium2007-vn}.
Statistics were largely unchanged by using a different map
(Figure~\ref{fig:supp-omni-hapmap-comparison}), although both maps were
inferred using array data, which is sparser than whole-genome sequencing data.
In subsequent analyses, we used data based on recombination rates fromthe Omni YRI map.

@@

Some statistics, such as $\E[D_i D_j]$, require at least two diploid samples
from a population to compute. Since we used a single Neanderthal sample, such
statistics for the Neandertal population were not used in the fit. By contrast,
there are statistics that only require a single sample in a given population to
estimate. These include cross-population nucleotide diversity measures, as well
as some LD statistics involving more than one population. For example,
statistics of the form $D_{human}(1-2p_{human})(1-2q_{neanderthal})$ require a
single Neanderthal sample and are informative of the Neanderthal demography.
These statistics were included in the fit, but statistics requiring more than
one Neanderthal sample to estimate were removed.

@@

The conditional site frequency spectrum (or cSFS) is the distribution of allele
frequencies restricted to loci that satisfy a given condition. Specifically, we
consider the distribution of allele frequencies in present-day populations
conditioned on the Vindija Neanderthal carrying the derived allele relative to
the inferred ancestral allele. Ancestral alleles alleles were determined from a
6 primate alignment \citep{1000_Genomes_Project_Consortium2015-zq}.
This cSFS is expected to be close to
uniform under neutrality and a simple split model (with no subsequent
migration) between the ancestors of humans and Neanderthals
\citep{Chen2007-iy}. By contrast, a U-shaped distribution has been taken as
evidence for introgression from an archaic hominin population that split from humans at least as far back in time as the human-Neanderthal split
\citet{Durvasula2020-td,Yang2012-ze}.
Sites with no calls in the Vindija Neanderthal were excluded from
this analysis.

@@

\section{Model specification and fitting}
\label{modelspec}
For the early history, we tested model parameterizations that cover many of the
proposed scenarios of population structure, size changes, and/or archaic
hominin admixture. The simplest model, in terms of number of parameters, was a
single-origin expansion of humans, with no structure in the stem and no
archaic hominin admixture aside from the Neanderthal admixture in Eurasian populations
following the out-of-Africa migration. This model allowed for a population size
change in the stem of humans between the ancestral split of the
human-Neanderthal lineages and the more recent split of branches leading to
Southern and West/Eastern African populations.

@@

From the reconstructed gene genealogies, we computed coalescence rates within
and between populations using \Relate’s function \texttt{RelateCoalescentRate}.
This allows for an estimate of the instantaneous inverse coalescence rate
(IICR) for samples drawn within each population (the inverse of which is often
interpreted as the effective population size history), and the relative cross
coalescence rates between pairs of populations (which are commonly used to
estimate divergence times).

@@

$F_{ST}$ represents the proportion of genetic variance in a multi-population
sample that can be attributed to drift between populations in that sample
(e.g., \cite{Bhatia2013-sq}).  $F_{ST}$ depends on population divergence times,
gene flow, and population sizes. For present-day sampled populations, predicted
$F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}), while past
estimates represent model predictions about the differentiation between past
populations. We computed $F_{ST}$ as a ratio of expected (or average)
variances, using observed pairwise diversity within and between populations as
in \citet{Peter2016-ys}:
\begin{equation}
    F_{ST}(i, j) = \frac{\E[2H_{i,j} - H_i - H_j]}{\E[2H_{i,j} + H_i + H_j]},
    \label{eq:FST}
\end{equation}

where $H_{ij}$ is the average number of differences between one haploid copy of
the genome from population $i$ and one from population $j$, and $H_i$ and $H_j$
are the average number of differences for two haploid copies from the same
population. In all inferred models, $F_{ST}$s between Neanderthals and the
ancestors of contemporary humans were much larger than between any pair of contemporary
human populations or their ancestors (stems). $F_{ST}$ between human
populations over the last $100ka$ remained relatively low, at $\sim0.1$ or
less. In the continuous-migration model, the two stem populations also were
predicted to have low $F_{ST}$ within this range, as migration rates were
inferred to be high enough to maintain genetic similarity. In both the
merger-without-stem-migration and the merger-with-stem-migration, one of the
stem populations was inferred to have a short, severe bottleneck, which led to
increased $F_{ST}$ between stem populations during that time. In the merger
with stem migration, $F_{ST}$ reached a maximum value of $0.3$ between stems,
intermediate between human-Neanderthal divergence ($0.45$)
and the maximum $F_{ST}$ observed between contemporary human populations
($\lessapprox 0.1$.)

@@

\subsection{\bm{$f_4$} statistics between pairs of contemporary
and pairs of ancient populations}
\label{sec:f4}

%\tdwcomment{I have tried to make the terminology about human, hominin, etc.
%consistent and unambiguous. To do this, I have avoided using "modern" because 
%"modern human" can mean "anatomically modern human" or \emph{Homo sapiens}, 
%but mostly in the paper we use it to mean "contemporary" or "present-day". I 
%think we should use "contemporary human" or "present-day human" when we mean 
%the latter and just "human" when we mean the former. I have made these changes 
%in the main text and supplement, preferentially using "contemporary" but 
%leaving "present-day" when it was already there. However, I think it would be 
%best if Simon made these changes in the equations in this section, so I don't mess 
%something up.}

@@

Recent studies have focused on characterizing the timing of the earliest
divergences among human populations, and such analyses
typically fit simple isolation-with-migration (IM) or ``clean split'' models to
observed allele frequencies or inferred coalescence times among and between
populations (see \citet{Weaver2008-ho,Bergstrom2021-iw} for recent reviews).
Simple IM models
typically fit a constant $N_e$ within the two diverged population, the split
time, and a symmetric continuous migration rate (which could be fixed to zero,
equivalent to the ``clean split'' model). We explored the effect of fitting
such simplified models of history to data from more complex models by
simulating the site frequency spectrum (SFS) under each of our four models for
pairs of populations and reinferring an IM and clean split model for each pair.
The simulations sampled 10 diploid individuals from each
population, and we used \msprime \citep{Baumdicker2022-mj} to simulated
500 Mb of total sequence length (the sum over500 replicates each 1 Mb in length)
with constant recombination rate $r=10^{-8}$ and mutation rate $u=2\times10^{-8}.$

>>>>

DNA samples were collected from three Nama communities in the Richtersveld
region of South Africa, which borders southern Namibia, in 2012. A community
guide was present during each interview and facilitated consent in Afrikaans or
Nama. Written consent was recorded per our IRB protocol with human subjects
approval from Stanford University (Protocol \#13829), Stellenbosch University
(N11/07/210) and later maintained via SUNY Stony Brook (Protocol \#727494).
Saliva samples were extracted from Oragene [OGR-500] kits at Stellenbosch
University. Results stemming from genetic analyses have been communicated in
2015, 2019 and 2021 via community presentations, a radio interview and to
representatives of the Richtersveld National Park.

@@

From the merged dataset of the African Diversity Reference Panel (ADRP)
\citep{Gurdasani2015-qy,Pagani2015-pz}
and the \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data,
we subsampled to the populations included in this study. From the 1000 Genomes
data, these included all populations in the African super-population (ESN, GWD,
LWK, MSL, and YRI) and four non-African populations (CEU, GBR, CHB, and PJL).
From the ADRP, these included the Amhara, Bakiga, Gumuz, Nama, Oromo, Somali,
Nama, and Zulu. This larger set of populations were used in the PCA and
\texttt{Admixture} analyses. The \texttt{Relate} analysis used a slightly
smaller set, discluding the Bakiga, Somali, and Zulu samples.

@@

All analyses presented in this work focus on biallelic single nucleotide
polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the
\texttt{moments-LD} analysis, we focused on intergenic regions because these
appear less affected by natural selection compared to both synonymous and
nonsynonymous variation \citep{Ragsdale2018-dd}, removing regions that fell
outside of the 1000 Genomes strict callability mask. To enable comparison with
Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
had less than 100 contiguous base pairs.

@@

We used principal component analysis (PCA) implemented in \texttt{scikit-learn}
\citep{Pedregosa2011-ke} and \texttt{ADMIXTURE} version 1.3.0
\citep{Alexander2011-aq} to summarize genetic diversity among a broader set of
1000 Genomes and ADRP populations (Figure~\ref{fig:diversity} in the main text). Data
filtering was identical for both analyses. We retained biallelic SNPs with no
missing data and minor allele frequency greater that 5\%. We then thinned the
remaining sites to remove SNPs in high linkage disequilibrium with one another.
We used the \texttt{locate\_unlinked} function from \texttt{scikit-allel}
version 1.2.1 \citep{Miles2021-yq} to identify SNPs in low linkage
disequilibrium, using a threshold of $r^2=0.1$, a window size of 100 SNPs, and
a step size of 20 SNPs, removing the remaining SNPs. This process was iterated
five times for each chromosome, resulting in 327,656 roughly unlinked SNPs used
in the PCA and \texttt{ADMIXTURE} analyses.

@@

Given two biallelic loci, with alleles $A$ and $a$ and the left locus and
alleles $B$ and $b$ at the right locus, the standard covariance measure of LD is
$D=p_{AB}p_{ab}-p_{Ab}p_{aB}$, where $p_{AB}$ is the frequency of $AB$
haplotypes in a population (and thus the probability of drawing an $AB$
haplotype in a random sample of that population). \citet{Hill1968-vu} solve for
the expectation of $D^2$ using a system of equations that includes $\E[Dz] =
\E[D(1-2p_A)(1-2p_B)]$ and $\E[\pi_2] = \E[p_A(1-p_A)p_B(1-p_B)]$, where $p_A$
and $p_B$ are the frequencies of $A$ and $B$ at the left and right loci,
respectively. This system also requires the expected pairwise diversity (or
expected heterozygosity, assuming random mating), denoted $\E[H] =
\E[2p_A(1-p_A)] = \E[2p_B(1-p_B)]$, assuming equal mutation rates at the two
loci.

@@

$\sigma_{Dz}$ has been shown to be sensitive to deep population structure and
archaic hominin admixture \citep{Ragsdale2019-nt}, and this statistic is closely
related to $S^*$ statistics used to scan for introgressed haplotypes
\citep{Plagnol2006-lt}. Pairwise diversity statistics $H_{i,j}$ have also been
widely used in demographic inference involving ancient DNA and many statistics, as
$f_2$, $f_3$ and $f_4$ statistics can be expressed as linear combinations of
$H_{i,j}$. $f$-statistics form the basis of admixture graph analysis
\citep{Lipson2020-cq}. Therefore, the set of statistics used here encompass
multiple features of genetic data that have been used to infer models of
archaic hominin admixture and population structure involving many populations.

@@

For numerical convenience, observed genetic variants were binned by
recombination distances. We assessed the robustness of the statistics to errors
in the recombination maps by considering two different recombination maps, the
OMNI YRI and HapMapII
\citep{1000_Genomes_Project_Consortium2015-zq,International_HapMap_Consortium2007-vn}.
Statistics were largely unchanged by using a different map
(Figure~\ref{fig:supp-omni-hapmap-comparison}), although both maps were
inferred using array data, which is sparser than whole-genome sequencing data.
In subsequent analyses, we used data based on recombination rates from the OMNI YRI map.

@@

Some statistics, such as $\E[D_i D_j]$, require at least two diploid samples
from a population to compute. Since we used a single Neanderthal sample, such
statistics for the Neandertal population were not used in the fit. By contrast,
there are statistics that only require a single sample in a given population to
estimate. These include cross-population nucleotide diversity measures, as well
as some LD statistics involving more than one population. For example,
statistics of the form
$D_{\text{human}}(1-2p_{\text{human}})(1-2q_{\text{neanderthal}})$ require a
single Neanderthal sample and are informative of the Neanderthal demography.
These statistics were included in the fit, but statistics requiring more than
one Neanderthal sample to estimate were removed.

@@

The conditional site frequency spectrum (or cSFS) is the distribution of allele
frequencies restricted to loci that satisfy a given condition. Specifically, we
consider the distribution of allele frequencies in present-day populations
conditioned on the Vindija Neanderthal carrying the derived allele relative to
the inferred ancestral allele. Ancestral alleles alleles were determined from a
6 primate alignment \citep{1000_Genomes_Project_Consortium2015-zq}.
This cSFS is expected to be close to
uniform under neutrality and a simple split model (with no subsequent
migration) between the ancestors of humans and Neanderthals
\citep{Chen2007-iy}. By contrast, a U-shaped distribution has been taken as
evidence for introgression from an archaic hominin population that split
from humans at least as far back in time as the human-Neanderthal split
\citep{Yang2012-ze,Durvasula2020-td}.
Sites with no calls in the Vindija Neanderthal were excluded from
this analysis.

@@

\section{Model specification and fitting}
\label{modelspec}
For the early history, we tested model parameterizations that cover many of the
proposed scenarios of population structure, size changes, and/or archaic
hominin admixture. The simplest model, in terms of number of parameters, was a
single-origin expansion of humans, with no structure in the stem and no
archaic hominin admixture aside from the Neanderthal admixture in Eurasian populations
following the out-of-Africa migration. This model allowed for a population size
change in the stem of humans between the ancestral split of the
human-Neanderthal lineages and the more recent split of branches leading to
southern and western/eastern African populations.

@@

From the reconstructed gene genealogies, we computed coalescence rates within
and between populations using \texttt{Relate}’s function \texttt{RelateCoalescentRate}.
This allows for an estimate of the instantaneous inverse coalescence rate
(IICR) for samples drawn within each population (the inverse of which is often
interpreted as the effective population size history), and the relative cross
coalescence rates between pairs of populations (which are commonly used to
estimate divergence times).

@@

$F_{ST}$ represents the proportion of genetic variance in a multi-population
sample that can be attributed to drift between populations in that sample
(e.g., \cite{Bhatia2013-sq}).  $F_{ST}$ depends on population divergence times,
gene flow, and population sizes. For present-day sampled populations, predicted
$F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}), while past
estimates represent model predictions about the differentiation between past
populations. We computed $F_{ST}$ as a ratio of expected (or average)
variances, using observed pairwise diversity within and between populations as
in \citet{Peter2016-ys}:
\begin{equation}
    F_{ST}(i, j) = \frac{\E[2H_{i,j} - H_i - H_j]}{\E[2H_{i,j} + H_i + H_j]},
    \label{eq:FST}
\end{equation}
where $H_{ij}$ is the average number of differences between one haploid copy of
the genome from population $i$ and one from population $j$, and $H_i$ and $H_j$
are the average number of differences for two haploid copies from the same
population.

In all inferred models, $F_{ST}$ between Neanderthals and the
ancestors of contemporary humans were much larger than between any pair of contemporary
human populations or their ancestors (stems). $F_{ST}$ between human
populations over the last $100ka$ remained relatively low, at $\sim0.1$ or
less. In the continuous-migration model, the two stem populations also were
predicted to have low $F_{ST}$ within this range, as migration rates were
inferred to be high enough to maintain genetic similarity. In both the
merger-without-stem-migration and the merger-with-stem-migration, one of the
stem populations was inferred to have a short, severe bottleneck, which led to
increased $F_{ST}$ between stem populations during that time. In the
merger-with-stem-migration, $F_{ST}$ reached a maximum value of $0.3$ between stems,
intermediate between human-Neanderthal divergence ($0.45$)
and the maximum $F_{ST}$ observed between contemporary human populations
($\lessapprox 0.1$.)

@@

\subsection{Shared differentiation between pairs of contemporary
and pairs of ancient populations}
\label{sec:f4}

@@

Recent studies have focused on characterizing the timing of the earliest
divergences among human populations, and such analyses
typically fit simple isolation-with-migration (IM) or ``clean split'' models to
observed allele frequencies or inferred coalescence times among and between
populations (see \citet{Weaver2008-ho,Bergstrom2021-iw} for recent reviews).
Simple IM models
typically fit a constant $N_e$ within the two diverged population, the split
time, and a symmetric continuous migration rate (which could be fixed to zero,
equivalent to the ``clean split'' model). We explored the effect of fitting
such simplified models of history to data from more complex models by
simulating the site frequency spectrum (SFS) under each of our four models for
pairs of populations and reinferring an IM and clean split model for each pair.
The simulations sampled 10 diploid individuals from each
population, and we used \msprime \citep{Baumdicker2022-mj} to simulated
500 Mb of total sequence length (the sum over 500 replicates each 1 Mb in length)
with constant recombination rate $r=10^{-8}$ and mutation rate $u=2\times10^{-8}.$
